---
title: "Assignment 3: Data Exploration"
author: "Abrão Soares Pereira"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup,  message = FALSE}
# Loading packages
library(tidyverse)
library(lubridate)
library(here)

# Get file path
getwd()

# Read or get file
Neonics <- read.csv(
  file = here("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"),
  stringsAsFactors = TRUE)  

# Read or get second file: Litter data
Litter <- read.csv(
  file = here("Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE)

```



## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Based on the dataset and information online, we are interested in the ecotoxicology of neonicotinoids because they are a  insecticide designed to kill pest insects in agriculture and gardens. However, their use can harmful consequences for other insects and bugs.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Studying forest litter is crucial because it's a key part of nutrient cycling. Which the litter would turn into organic matter that can help support soil health and other small organisms for growth. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Different trap and in sizes: litter is using elevated 0.5m2 PVC traps while fine woody
 debris is collected using ground traps. 
 2. Sample frequency: elevated traps are sampled every two weeks and ground traps are sampled
 per year. 
 3. Collected sample is shorted into functional groups and weight, masses are then reported per trap per collection event.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}

dim(Neonics)
dim(Litter)

# The dimensions of Neonics data is 4623 observations and 30 variables
# The dimensions of Litter data is 188 observations and 19 variables
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Q6}
effect_summary <- summary(Neonics)
sort(effect_summary, decreasing = TRUE)

effect_summary2 <- summary(Litter)
sort(effect_summary2, decreasing = TRUE)

# Create a summary of the "Effect" column
effect_summary <- summary(Neonics$Effect)
effect_summary <- summary(Litter$Effect)
sort(summary(Neonics$Effect), decreasing = TRUE)


```

> Answer: Based on the result, the most common effects that studied are population, mortality,
feeding behavior, reproduction and development


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Q7}

summary(Neonics$Species.Common.Name, maxsum = 6) 

species_summary <- summary(Neonics$Species.Common.Name)
species_table <- tibble::enframe(species_summary, name = "Species", value = "Count")

species_table <- species_table %>%
  arrange(desc(Count))

print(species_table)


```

> Answer: The six most common studied species in the dataset are honey bee, parasitic wasp,
buff tailed bumblebee, carniolan honey bee, bumble bee, Italian honey bee. Most of this top 6 species are important pollinators. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}
class(Neonics$Conc.1..Author.)
head(Neonics$Conc.1..Author.)
```

> Answer: The class is Factor, and its because the columns contains non-numeric characters and texts. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Q9}

ggplot(data = Neonics, aes(x = Publication.Year)) +
  geom_freqpoly(binwidth = 1) +
  labs(title = "Number of Studies Conducted by Publication Year",
       x = "Publication Year", y = "Number of Studies")

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Q10}

ggplot(data = Neonics, aes(x = Publication.Year, color = Test.Location)) +
  geom_freqpoly(binwidth = 1) +
  labs(title = "Number of Studies Conducted  by Test Location",
       x = "Publication Year", y = "Number of Studies")

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are in Lb, follow by field natural. Lab location is
increasing in trends by decades, and the field slowy increaing over the years with a big jump in the
middle 2000s but then drop drown in the middle of 2010 and 2020. I assume mostly due to funding or covid-19 influence. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}

ggplot(data = Neonics, aes(x = Endpoint)) +
  geom_bar() +
  labs(title = "Count of Study Endpoints", x = "Endpoint", y = "Count") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer: The two commons end points are LEOL and NOEL. Unless I misunderstood the question 
but the graph from my plot shows those tow are with the highest. However, LC50 and LD50 are both 
also have the highest end points. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}

class(Litter$collectDate)
Litter$collectDate <- ymd(Litter$collectDate)
class(Litter$collectDate)

# Find the unique dates
unique(Litter$collectDate)
```

13. Using the `unique` function, list the different plotIDs sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}

unique(Litter$plotID)

```

> Answer: The summary function return frequency in a table showing the most common plot IDs, while the unique function returns vector listings plot IDs that shows in the data. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}

ggplot(data = Litter, aes(x = functionalGroup)) +
  geom_bar() +
  labs(title = "Litter Counts Collected by Functional Group",
       x = "Functional Group", y = "Count")

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Q15}

ggplot(data = Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_boxplot() +
  labs(title = "dryMass of Litter by Functional Group")

# Violin plot 
ggplot(data = Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_violin()

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: Boxplot is easy to read and it shows the distribution clearly by showing median. Violin plot 
is viaully not effective and might be hard to interpret by general audience and its less informative. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles have the highest biomass and follows by twigs/branch. 
